---
title: "tree-based-methods"
author: "Ngoc Duong"
date: "4/20/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lasso2)
library(tidyverse)
library(ISLR)
library(caret)
library(rpart)
library(rpart.plot)
library(party)
library(partykit)
library(randomForest)
library(ranger)
library(gbm)
library(plotmo)
library(pdp)
library(lime)
```


### Problem 1

```{r}
data("Prostate") #import data
```

(i) Fit a regression tree with "lpsa" as the response and the other variables as predictors

```{r}
set.seed(13)
tree1 <-rpart(formula = lpsa~., data = Prostate)
rpart.plot(tree1)
```

Next, we can use cross-validation to determine the optimal tree.

```{r}
cpTable <-printcp(tree1)
plotcp(tree1)
```

We can prune the tree based on the cp table 

```{r}
minErr <-which.min(cpTable[,4]) # minimum cross-validation error
tree3 <-prune(tree1, cp = cpTable[minErr,1])# 1SE rule

tree4 <-prune(tree1, cp = cpTable[cpTable[,4]<cpTable[minErr,4]+cpTable[minErr,5],1][1])
```

The tree-size corresponding to the lowest cross-validation error is 7 (or 7 splits), compared to 3 splits in the tree obtained by the 1SE rule. 

(ii) Plot the final tree choice, and interpret a terminal node

```{r}
rpart.plot(tree3)
```

Let's pick terminal node ...

(iii) Perform bagging and report the variable importance 

```{r}
set.seed(13)
bagging <-randomForest(lpsa~., Prostate,mtry = 19)

set.seed(13)
rf <-randomForest(lpsa~., Prostate,mtry = 6)
# fast implementation

set.seed(13)
rf2 <-ranger(lpsa~., Prostate,mtry = 6)
# scale permutation importance by standard error
predict(rf, newdata = Hitters2[1:5,])
```


Get variable importance 
```{r}
set.seed(13)
rf2.final.imp <-ranger(Salary~., Hitters,
                       mtry = 3, 
                       splitrule = "variance",
                       min.node.size = 5,
                       importance = "impurity")

barplot(sort(ranger::importance(rf2.final.imp), decreasing = FALSE),
        las = 2, 
        horiz = TRUE, 
        cex.names = 0.7,
        col =colorRampPalette(colors =c("darkred","white","darkblue"))(19))
```





